---
title: "4 BenchmarkStudy"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{4 BenchmarkStudy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
--- 

```{r setup, warning=FALSE, message=FALSE, include=FALSE}
knitr::opts_chunk$set(warning = TRUE)
# devtools::load_all()
.libPaths("/dora/nobackup/yuec/R")
library(BenchHub)
library(R6)
library(scran)
```
  
```{r}
library(BenchHub)
```
  
  
# Overview

The `BenchHubStudy` object is designed to encapsulate all necessary components in a benchmarking study, including the data and functions associated. It provides a unified structure for benchmark developers to share their work and for method developers to interact with an existing benchmark study.     
 
- **Benchmark developers** can store *Trio* objects (containing the input data, metrics, and supporting evidence), any mapping functions and distribute a ready-to-use study object.     
- **Method developers**  can apply their methods to the provided data and evaluate their outputs using the built-in metrics.  

This vignette provides guide for both usage.  

 

# For Benchmark Developer  

This section demonstrates how to create a `BenchmarkStudy` object from a benchmarking study. 

## Initialising the Study

We begin by creating an empty `BenchmarkStudy` object.  

```{r}

study <- BenchmarkStudy$new()
 
```
 
## Adding Trio

Assume the benchmark developer has created a trio object for one of the data used for the benchmarking. Here, we use a Trio from the curated trio dataset. 

```{r}
# Download the Trio from the database using its name 
example_trio <- suppressMessages( Trio$new("benchhub_vignette_example", cachePath = TRUE) )

# Add to study
study$addTrio(example_trio)


```

## Define mapping function

A mapping function is a helper function that process the data into a format that can then be input into the evaluation metrics.    

For example, in simulation studies, the output from a simulator is a cell x gene count matrix. Suppose we want to compare the sparsity of genes between the simulated data and an experimental data, we need a mapping function that calculates the sparsity of genes from the count matrix.  

We define two examples of mapping functions.  

Example 1: calculate the sparsity (proportion of zero counts) per gene.  

```{r}
# Define the mapping function 
proportion_zero_gene <- function(data) {
  return(colMeans(counts(data) == 0))
}

# Add the mapping function
study$addMappingFunction(
  name = "Fraction Zero Genes",
  func = proportion_zero_gene,
  inputDescription = "SingleCellExperiment or matrix with gene expression counts.",
  outputDescription = "Numeric vector of fraction of zero counts per gene.",
  exampleUsage = paste(
    "## Minimal example",
    "#mat <- matrix(rpois(100, lambda = 1), nrow = 10, ncol = 10)",
    "#mat <- SingleCellExperiment(assays = list(counts = mat)",
    "#res <- study$runMapping('Fraction.Zero.Genes', mat)",
    "#head(res)",
    sep = "\n"
  )
)
 
 
```
 
Example 2: calculate the normalized library size per cell.  

```{r}

# Define the mapping function 
norm_lib_size <- function(data) {
  lib_size <- log1p(rowSums(counts(data)))
  dge <- edgeR::DGEList(counts = Matrix::t(counts(data)))
  norm_factors <- edgeR::calcNormFactors(dge, method = "TMM")$samples$norm.factors
  return(unname(lib_size * norm_factors))
}

# Add the mapping function, it is optional but recommended to add example usage 
study$addMappingFunction(
  name = "normalized library size",
  func = norm_lib_size,
  inputDescription = "SingleCellExperiment or matrix with gene expression counts.",
  outputDescription = "Numeric vector of normalised library size per cell."
)

```

 

## Uploading to Curated Trio Datasets


Once the BenchmarkStudy object includes:
- A study name and description    
- At least three Trio objects   
- Mapping functions [optional]      

The BenchmarkStudy object can be uploaded to the database. 

```{r}
# because we need at least three Trio 
# pretend we have added three unique Trios 
study$addTrio(example_trio) 
study$addTrio(example_trio)

# Note that each Trio needs to be already exist in the database. 
# Trios can be uploaded to the database using the `writeCTD()` function. 
# For more details, see the vignette on Trio construction.
# example_trio$writeCTD("example_trio")

# Set name and description manually
study$name <- "Benchhubstudy vignette"
study$description <- "This study compares simulated spatial transcriptomics data."

```

The object can be uploaded using the `writeBenchmarkStudy()` function.
```{r}
# We comment it out in the vignette so that it does not get repetitively added
# study$writeBenchmarkStudy()

```




 
# For Method Developer

This section illustrates how a method developer can use the benchmark study object created by another user, apply their method, and evaluate its performance.   

## Loading the Study

A BenchmarkStudy object can be downloaded from the database through its name.  

```{r message=FALSE}
study  <- suppressMessages( BenchmarkStudy$new("Benchhubstudy vignette", fetchFromCtd = T))
```



Inspect the list of available trios, and available mapping functions 

We see that this study has three Trios. Each Trio has supporting evidence that we can compare with. 

```{r}
length(study$trios)

study$trios[[1]]

```


This study provides two mapping function to process the data into a format that can be used for evaluation. 


Each mapping function has documentation.  

```{r}

study$listMappingFunctions()
doc <- study$getMappingFunctionDocumentation("Fraction.Zero.Genes" )
 
doc 
# Because the example Usage is a multiple line vector, need to use cat to print it nicely.
cat(doc$exampleUsage, "\n")
```

## Preparing for evaluation

This benchmark study wants to assess the quality of simulated data. 

Suppose the method developer has generated a simulated data with their method, and they want to compare with the real data to see how realistic the simulated data is.  

```{r}
# Suppose this is a simulated data 
set.seed(0)
sim <- mockSCE(ncells =  4744, ngenes = 1000)
```

The method developer can apply the mapping functions to the simulated data to generate the relevant features to evaluate such as the sparsity of the genes and the library size.  
 
```{r}

mydata_prop_zero_gene <- study$runMapping("Fraction.Zero.Genes", sim)
mydata_lib_size <- study$runMapping( "normalized.library.size", sim)
```

## Evaluate 


Now we can compare the simulated data against an experimental data using the `evaluate` function.

The evaluate function is in the format of `study$evaluate(trio_name, list( supporting evidence = output to compare with ))`. 

In the function below, the `Fraction zero genes` and `normalized library size` are the names of the supporting evidence that can be found in the `example_trio_1` Trio object. 


```{r}
 
result <- study$evaluate("benchhub_vignette_example",  # name of the trio to compare with  , can be accessed by study$trios[[1]]$name
  list("Fraction zero genes" = mydata_prop_zero_gene  , # name of the supporting evidence 
   "normalized library size" = mydata_prop_zero_gene )) # name of the supporting evidence   

# 
 
result


 
```

