---
title: "TrioR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TrioR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
--- 

```{r setup, message=FALSE}
devtools::load_all()
library(BenchHub)
```

# Import data

- Here, we use the survival data from survBenchmark-figShare. 
- This survival data is clinical (lung). 
- The Figshare ID is [26142922/47361046](https://figshare.com/articles/dataset/SurvBenchmark_comprehensive_benchmarking_study_of_survival_analysis_methods_using_both_omics_data_and_clinical_data/26142922?file=47361046) 
```{r}
testCache <- tempdir()
trio <- Trio$new("figshare:26142922/47361046", cachePath = testCache)
data <- trio |> purrr::pluck("data", 1)

dplyr::glimpse(data)
```

# Prepare survival model 

In this example, we use Cox proportional hazards regression model. 
Since we use c-index as evaluation metric, we here use one of the Cross-validation folds for an example. 

```{r}
cvSets <- cvTools::cvFolds(n = nrow(data), K = 3, R = 1)
j <- 1
test_id <- cvSets$subsets[cvSets$which == j]
test <- data[test_id, ]
train <- data[-test_id, ]
fit <- survival::coxph(survival::Surv(time, status) ~ age + sex + ph.ecog + pat.karno + meal.cal + wt.loss, data = train)
data$index <- ifelse(1:nrow(data) %in% test_id, "test", "train")
# all_surv_obj <- with(data, survival::Surv(time, status))
fit
```

# Preapre function

- surv_risk_pred: risk prediction/expectation
- surv_obj: survival object, this is for Gold Standard

```{r}
surv_risk_pred <- function(model, data) {
  if (is.null(data)) {
    return(unname(stats::predict(model, type = "risk")))
  } else {
    return(unname(stats::predict(model, newdata = data, type = "risk")))
  }
}
surv_obj <- function(data) {
  sub_data_test <- data[data[,"index"] == "test", ]
  sub_data_train <- data[data[,"index"] == "train", ]
  test_surv_obj <- with(sub_data_test, survival::Surv(time, status))
  train_surv_obj <- with(sub_data_train, survival::Surv(time, status))
  return(list(train_surv_obj, test_surv_obj))
}
```

Here are the metrics function

```{r}
harrelCIndexMetric <- function(to_eval, gs) {
  harrelC1 <- Hmisc::rcorr.cens(-to_eval[[2]], gs[[2]])
  return(harrelC1["C Index"])
}

beggCIndexMetric <- function(to_eval, gs) {
  begg_cindex <- survAUC::BeggC(gs[[1]], gs[[2]],to_eval[[1]], to_eval[[2]])
  return(begg_cindex)
}

unoCIndexMetric <- function(to_eval, gs) {
  # browser()
  uno_cindex <- survAUC::UnoC(gs[[1]], gs[[2]], to_eval[[2]])
  return(uno_cindex)
}

ghCIndexMetric <- function(to_eval, gs) {
  gh_cindex <- survAUC::GHCI(to_eval[[2]])
  return(gh_cindex)
}

brierScoreMetric <- function(to_eval, gs) {
  time <- gs[[1]][,"time"]
  brier_score <- survAUC::predErr(gs[[1]], gs[[2]],to_eval[[1]], to_eval[[2]],times=time, type = "brier", int.type = "unweighted")$error
  brier_score <- brier_score[is.finite(brier_score)]
  return(mean(brier_score))
}

timeDependentAUCMetric <- function(to_eval, gs) {
  time <- gs[[1]][,"time"]
  AUC_CD <- survAUC::AUC.uno(gs[[1]], gs[[2]], to_eval[[2]], time)
  return(mean(AUC_CD$auc))
}

```

# TrioR demo

```{r}
trio$data[[1]] <- data

expected <- list(surv_risk_pred(fit, NULL), surv_risk_pred(fit,test))

trio$addMetric("Harrel C-Index", harrelCIndexMetric)
trio$addMetric("Begg C-Index", beggCIndexMetric)
trio$addMetric("Uno C-Index", unoCIndexMetric)
trio$addMetric("GHCI C-Index", ghCIndexMetric)
trio$addMetric("Brier Score", brierScoreMetric)
trio$addMetric("Time Dependent AUC", timeDependentAUCMetric)

# surv_obj_actual <- surv_obj(data)
# trio$addGS("surv_obj", surv_obj_actual, c("Harrel C-Index", "Begg C-Index", "Uno C-Index", "GHCI C-Index", "Brier Score", "Time Dependent AUC"))
trio$addGS("surv_obj", surv_obj, c("Harrel C-Index", "Begg C-Index", "Uno C-Index", "GHCI C-Index", "Brier Score", "Time Dependent AUC"))
# trio$addGS("surv_obj", surv_obj, c("Harrel C-Index", "Begg C-Index", "Uno C-Index", "GHCI C-Index", "Brier Score", "Time Dependent AUC"))

actual <- trio$getGS("surv_obj") 
```

```{r}
evaluation <- trio$evaluate(list(surv_obj = expected))
evaluation
```

```{r}
harrelCIndexMetric(expected, actual[[1]]) == purrr::pluck(evaluation, "surv_obj", "Harrel C-Index")

beggCIndexMetric(expected, actual[[1]]) == purrr::pluck(evaluation, "surv_obj", "Begg C-Index")

unoCIndexMetric(expected, actual[[1]]) == purrr::pluck(evaluation, "surv_obj", "Uno C-Index")

ghCIndexMetric(expected, actual[[1]]) == purrr::pluck(evaluation, "surv_obj", "GHCI C-Index")

brierScoreMetric(expected, actual[[1]]) == purrr::pluck(evaluation, "surv_obj", "Brier Score")

timeDependentAUCMetric(expected, actual[[1]]) == purrr::pluck(evaluation, "surv_obj", "Time Dependent AUC")
```


Note:
- One TrioR only has one dataset
