---
title: "Introduction of Trio Class"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Overview of scFeatures with case studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
devtools::load_all()
library(BenchHub)
library(clusterProfiler)
library(DOSE)
library(DO.db)
library(EnsDb.Hsapiens.v86)
```

# Motivation

BenchHub—an R ecosystem to make benchmarking easier. It organizes evaluation metrics, gold standards(Auxiliary Data), and even provides built-in visualization tools to help interpret results. With BenchHub, researchers can quickly compare new methods, gain insights, and actually trust their benchmarking studies. 

# Creating Trio class

Trio objects can be created using the corresponding constructor. For example, if you have a dataset ID from Figshare, GEO, or ExperimentHub, you can create a Trio object as follows:

- figshare: Trio$new(figshare:figshareID)
- GEO: Trio$new(geo:GSEID)
- experiementhub: Trio$new(experiementhub:experimenthubID)

```{r, message=FALSE, warning=FALSE}
testCache <- tempdir()
trioA <- Trio$new("figshare:26142922/47361079", cachePath = testCache)
Trio$new("geo:GSE46474", cachePath = testCache)

getGEO
```

If you have your own dataset, you can easily create a trio object as well. Below is an example using a microbiome dataset.

```{r}
data("lubomski_microbiome_data", package = "BenchHub")
trioB <- Trio$new(data = lubomPD, datasetID = "lubomski_microbiome")
trioB
```

# Adding metric 

In benchmarking studies, a metric refers to the measurement used to evaluate a specific task. In this example, we define a task called survival model prediction.

As our evaluation metric, we use Harrel C Index statistics to assess the prediction.

First, we define a function, getHarrelCIndex, which computes Harrel C Index statistics:

```{r}
getHarrelCIndex <- function(to_eval, gs) {
  harrelC1 <- Hmisc::rcorr.cens(-to_eval[[2]], gs[[2]])
  return(harrelC1["C Index"])
}
```

After defining the metric function, we use the addMetric() function to incorporate it into the Trio object.

- The first argument specifies the metric name to be stored in the Trio object.
- The second argument is the function that defines the metric calculation.

In this case, we add the previously created getHarrelCIndex function as a metric named "Harrel C-Index":

```{r}
trioA$addMetric("Harrel C-Index", getHarrelCIndex)
```

# Adding AuxData

AuxData (Auxiliary Data) refers to gold standard in evaluation tasks and falls into two categories:

- Internal: Extracted directly from the Trio dataset. This may include metadata such as cell types (e.g., B cells, CD4, CD8), patient conditions (e.g., healthy, disease), and spatial domain information. The specific content depends on the dataset itself.
- External: Retrieved from external databases, such as disease pathway databases or gene marker databases.

## Internal

First, we define a function, getSurObj, which generate the survival object:

```{r}
getSurObj <- function(data) {
  sub_data_test <- data[data[,"index"] == "test", ]
  sub_data_train <- data[data[,"index"] == "train", ]
  test_surv_obj <- with(sub_data_test, survival::Surv(time, status))
  train_surv_obj <- with(sub_data_train, survival::Surv(time, status))
  return(list(train_surv_obj, test_surv_obj))
}
```

After defining the internal AuxData function, we use the addAuxData() function to incorporate it into the Trio object.

- The first argument specifies the AuxData name to be stored in the Trio object.
- The second argument is the function that defines the AuxData calculation.
- The third argument is the metric

In this case, we add the previously created getSurObj function as a AuxData named "survObj":

```{r}
trioA$addAuxData("survObj", getSurObj, c("Harrel C-Index"))
```

## External

This database contains a list of disease pathways, where each pathway is represented by a set of associated genes. The data was extracted from the Disease Ontology (DO) database and processed using the DOSE package to map disease terms to gene symbols. Once loaded, users can access the pathway information for further analysis, such as pathway enrichment or pathway activity scoring.

```{r}
DO_pathway <- load("~/Desktop/manuscript/TrioR/vignettes/DO_pathway.RData")
```

Next, we add this external database into trioA object. 

```{r}
trioA$addAuxData("getDOpathway", DO_pathway, c("diseaseScore"))
```

