---
title: "TrioR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TrioR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
--- 

```{r setup, message=FALSE}
devtools::load_all()
library(BenchHub)
library(limma)
```

# TrioB1

Calculate the accuracy of survival model prediction

- Here, we use the survival data from survBenchmark-figShare. 
- This survival data is clinical (lung). 
- The Figshare ID is [26142922/47361079](https://figshare.com/articles/dataset/SurvBenchmark_comprehensive_benchmarking_study_of_survival_analysis_methods_using_both_omics_data_and_clinical_data/26142922?file=47361079) 

```{r, message=FALSE}
testCache <- tempdir()
trioB1 <- Trio$new("figshare:26142922/47361079", cachePath = testCache)
data <- trioB1 |> purrr::pluck("data")
```

# Preprocessing: Feature selection of omics data

```{r}
data_filtered <- data[, !grepl("\\.\\.\\.1|time", colnames(data))]
data_filtered$status <- factor(data_filtered$status, levels = c(0, 1))
design <- model.matrix(~ 0 + data_filtered$status)
colnames(design) <- c("Group0", "Group1")

expression_data <- t(data_filtered[, sapply(data_filtered, is.numeric) & colnames(data_filtered) != "status"])

fit <- lmFit(expression_data, design)
contrast_matrix <- makeContrasts(Group1 - Group0, levels = design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

top_genes <- topTable(fit2, number = 100, adjust = "BH", sort.by = "P")

top_gene_names <- rownames(top_genes)
data_new <- data[, c(top_gene_names, "time", "status")]

trioB1$data <- data_new
```


Prepare survival model 

In this example, we use Cox proportional hazards regression model. 
Since we use c-index as evaluation metric, we here usethe Cross-validation folds for an example. 

Preapre function

- surv_risk_pred: risk prediction/expectation
- surv_obj: survival object, this is for Gold Standard

```{r}
surv_risk_pred <- function(model, data) {
  if (is.null(data)) {
    return(unname(stats::predict(model, type = "risk")))
  } else {
    return(unname(stats::predict(model, newdata = data, type = "risk")))
  }
}
surv_obj <- function(data) {
  sub_data_test <- data[data[,"index"] == "test", ]
  sub_data_train <- data[data[,"index"] == "train", ]
  test_surv_obj <- with(sub_data_test, survival::Surv(time, status))
  train_surv_obj <- with(sub_data_train, survival::Surv(time, status))
  return(list(train_surv_obj, test_surv_obj))
}

harrelCIndexMetric <- function(to_eval, gs) {
  harrelC1 <- Hmisc::rcorr.cens(-to_eval[[2]], gs[[2]])
  return(harrelC1["C Index"])
}
```

Add the metric and AuxData into the trioB1

```{r}
trioB1$addMetric("Harrel C-Index", harrelCIndexMetric)
trioB1$addAuxData("surv_obj", surv_obj, c("Harrel C-Index"))
```

# 5 repeated 5 fold cross validation

prepare cv index

```{r}
get_cv_index <- function(data, n, k) {
  
  all_repeats <- vector("list", length = n)
  
  for (r in 1:n) {  
    train_indices <- vector("list", length = k)
    test_indices <- vector("list", length = k)

    shuffled_indices <- sample(1:nrow(data))  
    fold_size <- ceiling(nrow(data) / k)
    folds <- split(shuffled_indices, ceiling(seq_along(shuffled_indices) / fold_size))

    for (i in seq_along(folds)) {
      test_indices[[i]] <- folds[[i]]  
      train_indices[[i]] <- setdiff(shuffled_indices, folds[[i]])  
    }
    
    all_repeats[[r]] <- list(train = train_indices, test = test_indices)
  }
  
  return(all_repeats)
}

```

prepare evaluation

```{r}
eval_repeated_cv_survival <- function(trio, cv_indices) {
  data <- trioB1 |> purrr::pluck("data")

  cv_repeats <- length(cv_indices) 
  cv_folds <- length(cv_indices[[1]]$train)  
  cindex_results <- numeric(cv_repeats * cv_folds)
  index <- 1

  for (r in seq_len(cv_repeats)) {
    train_indices_list <- cv_indices[[r]]$train
    test_indices_list <- cv_indices[[r]]$test
    
    for (j in seq_len(cv_folds)) {
      test_id <- test_indices_list[[j]]
      train_id <- train_indices_list[[j]]

      test <- data[test_id, ]
      train <- data[train_id, ]

      fit <- survival::coxph(survival::Surv(time, status) ~ ., data = train)

      data$index <- ifelse(1:nrow(data) %in% test_id, "test", "train")
      trioB1$data <- data

      predicted <- list(
        ifelse(is.infinite(surv_risk_pred(fit, NULL)), 0, surv_risk_pred(fit, NULL)),
        ifelse(is.infinite(surv_risk_pred(fit, test)), 0, surv_risk_pred(fit, test))
      )

      evaluation <- trioB1$evaluate(list(surv_obj = predicted))
      cindex_results[index] <- evaluation$surv_obj$`Harrel C-Index`
      index <- index + 1
    }
  }
  
  return(cindex_results)
}


```

sample case

```{r, message=FALSE, warning=FALSE}
set.seed(3)
data <- trioB1 |> purrr::pluck("data")
cv_indices <- get_cv_index(data, n = 5, k = 5)
cindex_results <- eval_repeated_cv_survival(trioB1, cv_indices)
```

Visualization

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
cindex_results <- data.frame(cindex_values = cindex_results)

th <-   theme(text=element_text(size=12 ),
              axis.text.x = element_text(angle = 45, hjust = 1),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              panel.background = element_rect(colour = "black", size=0.2, fill=NA) )

ggplot(cindex_results, aes(y = cindex_values)) +
  geom_boxplot() +
  labs(y = "C-Index") +
  theme_minimal() +
  th
```

# trioA

```{r, message=FALSE}
library(clusterProfiler)
library(DOSE)
library(DO.db)
library(EnsDb.Hsapiens.v86)
```

```{r}
get_DO_data <- function(ont="DO") {
  ont="DO"
  if (ont == "DO"){
      tryCatch(utils::data(list="DO2ALLEG", package="DOSE"))
      assign("DO2ALLEG", DO2ALLEG )
      DO2ALLEG <- get("DO2ALLEG")

      tryCatch(utils::data(list="EG2ALLDO", package="DOSE"))
      assign("EG2ALLDO", EG2ALLDO )
      EG2ALLDO <- get("EG2ALLDO")
    
    PATHID2EXTID <- get("DO2ALLEG" )
    EXTID2PATHID <- get("EG2ALLDO" )
    
    PATH2NAME.df <- toTable(DOTERM)
    PATH2NAME.df <- PATH2NAME.df[, c("do_id", "Term")]
    PATH2NAME.df <- unique(PATH2NAME.df)
    PATH2NAME <- PATH2NAME.df[,2]
    names(PATH2NAME) <- PATH2NAME.df[,1]
 
  }else if (ont == "DOLite"){
    
      tryCatch(utils::data(list="DOLite2EG", package="DOSE"))
      assign("DOLite2EG", DOLite2EG ) 
      DOLite2EG <- get("DOLite2EG")
      
      tryCatch(utils::data(list="EG2DOLite", package="DOSE"))
      assign("EG2DOLite", EG2DOLite )
      EG2DOLite <- get("EG2DOLite")
      
      tryCatch(utils::data(list="DOLiteTerm", package="DOSE"))
      assign("DOLiteTerm", DOLiteTerm )
      DOLiteTerm <- get("DOLiteTerm")
    
    PATHID2EXTID <- get("DOLite2EG")
    EXTID2PATHID <- get("EG2DOLite")
    PATH2NAME <- get("DOLiteTerm")
  }
   
  return( list(PATHID2EXTID = PATHID2EXTID , 
               PATH2NAME  = PATH2NAME 
               ) )
}

get_DO_pathway <- function(ont = "DO", patterns = c("diabetes", "cardiovascular", "heart", "cardiac", "hypertension")) {
  
  database <- get_DO_data(ont = ont)
  DOpathway <- database$PATHID2EXTID
  pathway_name <- database$PATH2NAME
  
  selected_pathway <- pathway_name[grep(paste0(patterns, collapse = "|"), pathway_name)]
  selected_pathway <- selected_pathway[!selected_pathway %in% c("carcinoma", "tumor", "cancer", "defect")]
  
  DOpathway <- DOpathway[names(DOpathway) %in% names(selected_pathway)]
  selected_pathway <- selected_pathway[match(names(DOpathway), names(selected_pathway))]
  names(DOpathway) <- unname(selected_pathway)
  
  DOpathway <- DOpathway[sapply(DOpathway, length) >= 10]
  
  for (i in seq_along(DOpathway)) {
    this <- DOpathway[i]
    getgene <- ensembldb::select(EnsDb.Hsapiens.v86,
                                 keys = unname(this)[[1]],
                                 keytype = "ENTREZID",
                                 columns = c("SYMBOL", "ENTREZID"))
    DOpathway[[i]] <- unique(getgene$SYMBOL)
  }
  
  return(DOpathway)
}

```

```{r, message=FALSE}
testCache <- tempdir()
trioA <- Trio$new("figshare:26142922/47361079", cachePath = testCache)
data <- trioA |> purrr::pluck("data")
data_log_normalized <- data %>%
  mutate(across(where(is.numeric), ~ log(. + 1)))
data <- data_log_normalized
```

```{r}
calculate_scores <- function(to_eval, gs) {
  colnames_data <- colnames(to_eval)
  scores <- sapply(names(gs), function(pathway) {
    genes <- intersect(gs[[pathway]], colnames_data)
    gene_means <- sapply(genes, function(gene) mean(to_eval[[gene]], na.rm = TRUE))
    sum(gene_means, na.rm = TRUE)
  })
  return(scores)
}

```

```{r}
trioA$addMetric("disease score", calculate_scores)
trioA$addAuxData("get_DO_pathway", get_DO_pathway, c("disease score"))
trioA$data <- data
# result <- calculate_scores(data, DOpathway)

evaluation <- trioA$evaluate(list(get_DO_pathway = data))
disease_scores <- evaluation$get_DO_pathway$`disease score`


df <- data.frame(
  Disease = names(disease_scores),
  Normalized_Score = disease_scores
)


ggplot(df, aes(x = reorder(Disease, Normalized_Score), y = Normalized_Score)) +
  geom_bar(stat = "identity", fill = "#E64525") +
  coord_flip() +  # Flip coordinates for better readability
  labs(x = "Disease", y = "Score") +
  th
```



