---
title: "How to create and use a Trio object"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to create and use a Trio object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
devtools::load_all()
```

# Initialising a Trio object

## Method 1: Curated Trio Datasets

You can directly use the name from the Curated Trio Datasets sheet to intialise a Trio object populated with some metrics and auxiliary data.

```{r}
trio <- Trio$new("Veteran_data", cachePath = tempdir())

trio
```

## Method 2: Source and ID

Trio objects can be created by specifying an ID from a source with a valid trio downloader.

Here, a Trio is initialised with a specific file from a figshare collection.

```{r}
trio <- Trio$new("figshare:26054188/47112109", cachePath = tempdir())

trio
```

## Method 3: Load an object directly

Trio can also be created by passing an object directly into the constructor. 

```{r}
data("lubomski_microbiome_data", package = "BenchHub")

trio <- Trio$new(data = lubomPD, datasetID = "lubomski_microbiome")

trio
```

## Bonus: Using a custom loader

Trio supports custom loaders for data formats not directly supported by Trio.

A loader is any function that takes in a path, provided by a downloader, and returns an object to be loaded into Trio.

Below, we use an anonymous function to wrap `GEOquery::getGEO` and provide it with the path of the downloaded file.

```{r}
trio <- Trio$new(
  "GEO:GSE46474",
  dataLoader = \(x) suppressMessages(GEOquery::getGEO(filename = x)),
  cachePath = tempdir()
)

trio
```


# Adding components to Trio

## Adding metrics

Adding metrics is simple. A metric is any pairwise function of the form `f(expected, predicted)` which returns a single value. 


```{r}
eq <- \(expected, predicted, inequality = FALSE) {
  if (inequality) {
    return(!expected == predicted)
  }

  expected == predicted
}

trio$addMetric("equality", eq)


# trio also supports passing through arguments to a metric
trio$addMetric("inequality", eq, args = list(inequality = TRUE))
```

## Adding and getting auxData

Auxilliary data is any data that allows you to make an evaluation.

Auxilliary data is added linked to metrics that allow you to evaluate it.

```{r}
# it can be a fixed quantity
trio$addAuxData(
  name = "Number One", auxData = 1, metrics = c("equality", "inequality")
)

# or it can be a quantity, to be computed from that dataset

trio$addAuxData(
  name = "length", auxData = length, metrics = c("equality", "inequality")
)
```

To view the auxiliary data one can use `getAuxData`.


```{r}
trio$getAuxData("length") # this is equivalent to nrow(trio$data)
```


# Evaluation

Once Trio is setup with relevant metrics and auxData, evaluation is trivial.

For simple evaluation:
```{r}
trio$evaluate(list("Number One" = 2, length = 182))
```

For multi-method evaluation.

```{r}
trio$evaluate(list(
  method1 = list("Number One" = 2, length = 182),
  method1 = list("Number One" = 1, length = 500)
))
```
