---
title: "How to create and use a Trio object"
date: "`r BiocStyle::doc_date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How to create and use a Trio object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r setup}
devtools::load_all()
```

# Initialising a Trio object

## Method 1: Curated Trio Datasets

You can directly use the name from the Curated Trio Datasets sheet to initialise a Trio object populated with some metrics and auxiliary data. This method is useful when you want to quickly start with a predefined dataset.

```{r}
trio <- Trio$new("Veteran_data", cachePath = tempdir())

trio
```

## Method 2: Source and ID

Trio objects can be created by specifying an ID from a source with a valid trio downloader. This method is useful when you have a specific dataset ID from a supported source like Figshare, GEO, or ExperimentHub.

Here, a Trio is initialised with a specific file from a Figshare collection.

```{r}
trio <- Trio$new("figshare:26054188/47112109", cachePath = tempdir())

trio
```

## Method 3: Load an object directly

Trio can also be created by passing an object directly into the constructor. This method is useful when you already have a dataset loaded in your R environment and want to use it with Trio.

```{r}
data("lubomski_microbiome_data", package = "BenchHub")

trio <- Trio$new(data = lubomPD, datasetID = "lubomski_microbiome")

trio
```

## Bonus: Using a custom loader

Trio supports custom loaders for data formats not directly supported by Trio. A loader is any function that takes in a path, provided by a downloader, and returns an object to be loaded into Trio.

Below, we use an anonymous function to wrap `GEOquery::getGEO` and provide it with the path of the downloaded file.

```{r}
trioGEO <- Trio$new(
  "GEO:GSE46474",
  dataLoader = \(path) suppressMessages(GEOquery::getGEO(filename = path)),
  cachePath = tempdir()
)

trioGEO
```


# Adding components to Trio

## Adding metrics

Adding metrics is simple. A metric is any pairwise function of the form `f(expected, predicted)` which returns a single value. Metrics are used to evaluate the performance of your methods against the gold standard.

```{r}
eq <- \(expected, predicted, inequality = FALSE) {
  if (inequality) {
    return(!expected == predicted)
  }

  expected == predicted
}

trio$addMetric("equality", eq)

# Trio also supports passing through arguments to a metric
trio$addMetric("inequality", eq, args = list(inequality = TRUE))
```

## Adding and getting auxData

Auxiliary data is any data that allows you to make an evaluation. Auxiliary data is added linked to metrics that allow you to evaluate it.

```{r}
# It can be a fixed quantity
trio$addAuxData(
  name = "Number One", auxData = 1, metrics = c("equality", "inequality")
)

# Or it can be a quantity, to be computed from that dataset
trio$addAuxData(
  name = "length", auxData = length, metrics = c("equality", "inequality")
)
```

To view the auxiliary data one can use `getAuxData`.

```{r}
trio$getAuxData("length") # This is equivalent to nrow(trio$data)
```


# Evaluation

Once Trio is set up with relevant metrics and auxData, evaluation is trivial.

For simple evaluation:
```{r}
trio$evaluate(list("Number One" = 1, length = 182))
```

For multi-method evaluation:
```{r}
trio$evaluate(list(
  great_method = list("Number One" = 1, length = 575),
  poor_method = list("Number One" = 2, length = 182)
))
```
