---
title: "1 Introduction to the Trio Class"
date: "`r BiocStyle::doc_date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{1 Introduction to the Trio Class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE
)
```

```{r, message=FALSE, warning=FALSE}
# devtools::load_all()
library(BenchHub)
library(clusterProfiler)
library(DOSE)
library(DO.db)
library(EnsDb.Hsapiens.v86)
```

# Motivation

BenchHub is an R ecosystem built to make benchmarking easier.

It organizes data, evaluation metrics, gold standards (auxiliary data), and even provides built-in visualization tools to help interpret results. With BenchHub, researchers can quickly compare new methods, gain insights, and produce trustworthy their benchmarking studies. 

# Creating Trio object

Trio objects can be created using the `Trio$new` constructor. There are 3 ways to create a Trio object:

- Curated Trio Datasets
- Source and ID
- Load an object directly

If the dataset can't be loaded using Trio's inbuilt loader, a custom loader can be provided.

## Method 1: Curated Trio Datasets

You can directly use the name from the Curated Trio Datasets sheet to initialise a Trio object populated with some metrics and auxiliary data. This method is useful when you want to quickly start with a predefined dataset.

```{r}
tempCache <- tempdir()
trio <- Trio$new("Veteran_data", cachePath = tempCache)

trio
```

The above output shows that we have a Trio with a dataset, metrics, and auxiliary data. The dataset contains 137 rows and 9 columns, and the metrics and auxiliary data are already populated and printed.

This Trio is ready for use in survival prediction evaluation.

## Method 2: Source and ID

Trio objects can be created by specifying an ID from a source with a valid trio downloader. This method is useful when you have a specific dataset ID from a supported source like Figshare, GEO, or ExperimentHub.

For example, if you have a dataset ID from Figshare, GEO, or ExperimentHub, you can create a Trio object as follows:

- figshare: `Trio$new("figshare:figshareID[/fileID]")`
  - `fileID` can optionally be provided to specify a specific file in the collection.
- GEO: `Trio$new("geo:GSEID[/Supplementary_filename]")`
  - `Supplementary_filename` can optionally be provided to specify a specific supplementary file in the series.
- experiementhub: `Trio$new("experiementhub:experimenthubID")`

The example below shows how to create a Trio object using a Figshare dataset with a `fileID`.

```{r, message=FALSE, warning=FALSE}
trioA <- Trio$new("figshare:26142922/47361079", cachePath = tempCache)

trioA
```

## Method 3: Load an object directly

Trio can also be created by passing an object directly into the constructor. This method is useful when you already have a dataset loaded in your R environment and want to use it with Trio.

If you have your own dataset, you can easily create a trio object as well. Below is an example using a microbiome dataset.

```{r}
data("lubomski_microbiome_data", package = "BenchHub")
trioB <- Trio$new(data = lubomPD, datasetID = "lubomski_microbiome")
trioB
```

## Bonus: Using a custom loader

Trio supports custom loaders for data formats not directly supported by Trio. A loader is any function that takes in a path, provided by a downloader, and returns an object to be loaded into Trio.

Below, we use an anonymous function to wrap `GEOquery::getGEO` and provide it with the path of the downloaded file.

```{r, eval=FALSE}
# This code will not run as it requires a GEO dataset to be downloaded
# GEO is unreliable
trioGEO <- Trio$new(
  "GEO:GSE46474",
  dataLoader = \(path) suppressMessages(GEOquery::getGEO(filename = path)),
  cachePath = tempdir()
)

trioGEO
```

# Adding components to Trio

## Adding metrics

In benchmarking studies, a metric refers to the measurement used to evaluate a specific task. In this example, we define a task called survival model prediction.

In Trio, a metric is any pairwise function of the form `f(expected, predicted)` which returns a single value.

We can also add metrics that have additional arguments by passing a list of arguments to the args parameter.

```{r}
eq <- \(expected, predicted, inequality = FALSE) {
  if (inequality) {
    return(!expected == predicted)
  }

  expected == predicted
}

trio$addMetric("equality", eq)

# Trio also supports passing through arguments to a metric
# Note: parameter names added for clarity
trio$addMetric(
  name = "inequality", metric = eq, args = list(inequality = TRUE)
)
```

In the above example, we added two metrics based on the same function: "equality" and "inequality". The "equality" metric checks if the expected and predicted values are equal, while the "inequality" metric checks if they are not equal.

Underneeth the hood, Trio creates a wrapper function that calls the metric function with the specified arguments.

```{r}
trio$metrics$inequality
```

## Adding and getting `auxData`

`auxData` (Auxiliary Data) refers to gold standard in evaluation tasks and falls into two categories:

- Internal: Extracted directly from the Trio dataset. This may include metadata such as cell types (e.g., B cells, CD4, CD8), patient conditions (e.g., healthy, disease), and spatial domain information. The specific content depends on the dataset itself.
- External: Retrieved from external databases, such as disease pathway databases or gene marker databases.

With the `equality` and `inequality` metrics added above, we can now run an evaluation.

First we add some auxiliary data to the Trio object to evaluate against.

Auxiliary data is any data that allows you to make an evaluation. Auxiliary data is added linked to metrics that allow you to evaluate it.

```{r}
# It can be a fixed quantity
trio$addAuxData(
  name = "Number One", auxData = 1, metrics = c("equality", "inequality")
)

# Or it can be a function that acts on the data
trio$addAuxData(
  name = "length", auxData = length, metrics = c("equality", "inequality")
)
```

To view the auxiliary data one can use `getAuxData`.

```{r}
trio$getAuxData("length") # This is equivalent to `length(trio$data)`
```


## Evaluation Example

Once Trio is set up with relevant metrics and auxData, we are ready to evaluate.

For simple evaluation:
```{r}
result <- trio$evaluate(list("Number One" = 1, length = 182))
knitr::kable(result)
```

For multi-method evaluation:
```{r}
result <- trio$evaluate(list(
  great_method = list("Number One" = 1, length = 575),
  poor_method = list("Number One" = 2, length = 182)
))
knitr::kable(result)
```

For a more detailed look at evaluation using Trio, please refer to vignette 2.


# Other Features 

## Caching

Trio uses caching to avoid lengthy downloads after the first time a data set is accessed. The `cachePath` parameter specifies the path to the cache directory. If not specified, the cache directory defaults to `~/.cache/R/TrioR/`.

## Data Splitting

Trio supports data splitting for cross-validation. The `split` method splits the data into training and test sets for cross-validation. The `splitIndices` attribute stores the indices for each sample.

Indices are generated using the `splitTools` package. The `split` method takes in the outcome variable and the number of folds and repeats. The `stratify` parameter can be used to stratify the outcome variable.

```{r}
trio$split(y = 1:137, n_fold = 2, n_repeat = 5, seed = 1234, stratify = FALSE)
trio$splitIndices
```

# Conclusion

In this vignette, we introduced the Trio class and demonstrated how to create a Trio object using curated datasets, source and ID, or loading an object directly. We also showed how to add metrics and auxiliary data to a Trio object and evaluate the performance of different methods using these metrics and auxiliary data. We hope this vignette helps you get started with Trio and conduct benchmarking studies more effectively.

# Session Info

```{r}
sessionInfo()
```
