---
title: "2 Evaluation using Trio"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{2 Evaluation using Trio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
library(BenchHub)
library(tidyverse)
library(glmnet)
```

# Import microbiome data

The `Trio` object in `BenchHub` can take datasets provided by users. To demonstrate, its ability to take user-provided datasets, we'll be using a microbiome dataset called `Lubomski` obtained from the `PD16Sdata` package. The following code will import the `Lubomksi` data into R. `lubomski_microbiome_data.Rdata` contains two data objects: `x` and `lubomPD`. `x` is a 575 by 1192 matrix containing the abundance of 1192 microbial taxa for 575 samples. `lubom_pd` is a factor vector of binary patient classes for 575 samples where where `1` represents `PD` and `0` represents `HC`.

```{r}
# import the microbiome data
data("lubomski_microbiome_data", package = "BenchHub")

# check the dimension of the microbiome matrix
dim(x)

# check the length of the patient status
length(lubomPD)
```

The task we'll be evaluating uses a binary classification task where each sample is either a Parkinson's Disease (PD) patient or Healthy Control (HC). Once the data are ready to be inputted to `Trio`, we can load `Trio`.

# Initialise a Trio object

To initialise a Trio object, we use a `new()` method.

```{r}
trio <- Trio$new(data = x, 
                 evidence = list(
                   Diagnosis = list(
                     evidence = lubomPD, 
                     metrics = "Balanced Accuracy"
                     )
                   ),
                 metrics = list(
                   `Balanced Accuracy` = balAccMetric
                   ),
                 datasetID = "lubomski_microbiome"
                 )
```

The supporting evidence is the disease diagnosis, which is the `lubomPD` vector that we extracted from the `Lubomski` data above and the relevant metric to evaluate it is Balanced Accuracy. Next, we will use this microbiome data to illustrate two examples, 1) evaluation with cross validation, 2) evaluation without cross validation.

# Cross validation - using the trio split function

In this example, we show a typical classification scenario to classify the patient status, where we need to split the data into training and testing set.

Step 1: Obtain relevant data for the model. To build a model, the following code will extract the data matrix and patient status outcome from `Trio` object.

```{r}
# get the gold standard from the Trio object
x <- trio$data
y <- trio$getEvidence("Diagnosis")
```

Step 2: For repeated cross-validation, the `split()` function will split the `y` vector (i.e., `evidence`) into the number of folds and repeats users want. In this case we used `n_fold = 2` and `n_repeat = 5` (i.e., 2-fold cross-validation with 10 repeats). Then users can get cross-validation indices (`CVindices`) for each sample, through `splitIndices`. This gives a simple list where each element represents a combination of folds and repeats for each sample.

```{r}
# get train and test indices
trio$split(y = y, n_fold = 2, n_repeat = 5)
CVindices <- trio$splitIndices
```

Step 5: Build the classification model and evaluate. We can now use a for loop to cross-validate evaluation results.

Using the `CVindices`, user can subset to training and test data. As an example, we build a LASSO regression model as a classification model on the training data, then make predictions on the test data.

Once predictions are obtained for the test set, we pass them to Trio using the same evidence name as stored in the Trio object (i.e., `Diagnosis`). Specifically, we call `trio$evaluate(list(lasso = list(Diagnosis = pred)))`, which instructs `evaluate()` to compare `pred` against the reference labels `Diagnosis` stored in `trio`, and then compute the specified metric (Balanced Accuracy).

```{r}
set.seed(1234)

# Loop through 2 folds x 5 repeats = 10 runs
result <- do.call(
  rbind,
  mapply(
    function(trainIDs, crossValID) {
      x_train <- x[trainIDs, ]
      x_test  <- x[-trainIDs, ]
      y_train <- y[trainIDs]
      y_test  <- y[-trainIDs]

      # Find the best lambda for LASSO regression
      cv_lasso <- cv.glmnet(
        x = as.matrix(x_train),
        y = y_train,
        alpha = 1,
        family = "binomial"
      )
      lam <- cv_lasso$lambda.1se

      # Fit a model with the best lambda on training data
      fit <- glmnet(
        x = as.matrix(x_train),
        y = y_train,
        alpha = 1,
        lambda = lam,
        family = "binomial"
      )

      # Evaluate the model on test data
      pred <- predict(
        fit,
        newx = as.matrix(x_test),
        s = lam,
        type = "class"
      )
      pred <- setNames(as.factor(as.vector(pred)), rownames(x_test))

      # Get the chosen evaluation metric from the Trio
      eval_res <- trio$evaluate(list(lasso = list(Diagnosis = pred)))

      # Keep track of the repeat and fold information
      eval_res$track <- crossValID
      eval_res
    },
    CVindices,
    names(CVindices),
    SIMPLIFY = FALSE
  )
)

```

After cross-validation, we can visualise cross-validation results by averaging results across folds within each repeats.

```{r fig.cap = "Mean cross-validation accuracy across repeats."}
result$fold <- unlist(lapply(strsplit(result$track, ".", fixed = TRUE), `[`, 1))
result$repeats <- unlist(lapply(strsplit(result$track, ".", fixed = TRUE), `[`, 2))

result <- result %>%
  dplyr::group_by(datasetID, method, evidence, metric, repeats) %>%
  dplyr::summarize(result = mean(result))

# visualise the result
boxplot(
  result$result,
  ylab = "Accuracy",
  main = "Cross-validation performance"
)
```

# Without Cross-validation

In this example, we show a simpler case without cross validation, where we have a prediction that we want to compare with the ground truth directly. For example, we may have simulated a single-cell count matrix and want to compare with an experimental count matrix on various attributes.

Here, we use the microbiome dataset to demonstrate the above case.

Step 1: Simulate a matrix of the same size as the microbiome data using negative binomial. Step 2: Define a function that calculate the sparsity of a given matrix. Step 3: Define a metric that compare difference of two values. Step 4: Evaluate the difference in sparsity of the microbiome data and the simulated data.

```{r}
set.seed(1)

# generate a simulated matrix
sim <- rnbinom(nrow(x) * ncol(x), size = 1, mu = 1)
sim <- Matrix(sim, nrow = nrow(x), ncol = ncol(x))

# function that calculate the sparsity of a matrix
calc_sparsity <- function(data) {
  sparsity <- sum(data == 0) / length(data)
  return(sparsity)
}


# metric that compare the difference of two values
calc_diff <- function(evidence, predicted) {
  return(predicted - evidence)
}
```

```{r}
# add metric that we just defined
trio$addMetric(name = "Difference", metric = calc_diff)

# calculate the sparsity of the data, and then input into the supporting evidence
trio$addEvidence(name = "Sparsity", evidence = calc_sparsity, metrics = "Difference")

# because we used negative binomial to simulate a matrix
# we will name the method as "negative binomial"
# then calculate the sparsity of the simulated matrix 
# and compare with the Sparsity supporting evidence
eval_res <- trio$evaluate(
  list(negative_binomial = list(Sparsity = sim))
)

eval_res
```

From the evaluation result, we see there is a `r round(eval_res$result, 2)` difference in the sparsity between the microbiome data and the data simulated with negative binomial.

# Importing result to BenchmarkInsights

The result dataframe obtained from `Trio` evaluation can be passed to `BenchmarkInsights` for subsequent visualisation. `BenchmarkInsights` is another structure in `BenchHub` that provide a list of functions to analyse and visualise the benchmarking results from multiple perspectives.

Note, to make the result dataframe compatible for BenchmarkInsights, please make sure the dataframe is in the format shown below, where the column names are "datasetID", "method", "evidence", "metric" , "result", and that there is one row for each result.

```{r}
# in the with cross validation result, we need to average the results from multiple repeats to give one value
result <- result %>%
  dplyr::group_by(datasetID, method, evidence, metric) %>%
  dplyr::summarize(result = mean(result))

result <- rbind(result, eval_res)
result
```

Please see Vignette 3 for the details on the visualisations and functions in `BenchmarkInsights`.

# Session Info

```{r}
sessionInfo()
```
