---
title: "2 Evaluation using TrioR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TrioR_evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
--- 

```{r setup, message=FALSE}
devtools::load_all()
library(BenchHub)
library(tidyverse)
library(glmnet)
```

# Import microbiome data

`TrioR` can take datasets provided by users. To demonstrate, its ability to take user-provided datasets, we'll be using a microbiome dataset called `Lubomski` obtained from the `PD16Sdata` package. The following code will import the `Lubomksi` data into R. `lubomski_microbiome_data.Rdata` contains two data objects: `x` and `lubomPD`. `x` is a 575 by 1192 matrix containing the abundance of 1192 microbial taxa for 575 samples. `lubom_pd` is a factor vector of binary patient classes for 575 samples where where `1` represents `PD` and `0` represents `HC`.

```{r}
# import the microbiome data
data("lubomski_microbiome_data", package = "BenchHub")

# check the dimension of the microbiome matrix
dim(x)

# check the length of the patient status
length(lubomPD)
```

The task we'll be evaluating using `TrioR` is a binary classification task where each sample is either a Parkinson's Disease (PD) patient or Healthy Control (HC). Once the data are ready to be inputted to `TrioR`, we can load `TrioR`.


# Initialise a TrioR object

To initialise a TrioR object, we use a `new()` method.

```{r}
trio <- Trio$new(data = x, datasetID = "lubomski_microbiome")
```

Next we will use this microbiome data to illustrate two examples, 1) evaluation with cross validation, 2) evaluation without cross validation.     


# Cross validation - using the trioR split function

In this example, we show a typical classification scenario to classify the patient status, where we need to split the data into training and testing set.  

## Step 1: Add a metric to it using `Trio$addMetric()`

Users can use existing evaluation functions by specifying `metric`, and can name their evaluation metric by specifying `name`. Since the evaluation task is a classification task, we chose balanced accuracy, using the existing function bundled in BenchHub, and named it `Balanced Accuracy`.

```{r}
# add a metric to the Trio object
trio$addMetric(name = "Balanced Accuracy", metric = balAccMetric)
```

## Step 2: Add auxiliary data to the `TrioR` object using `Trio$addAuxData`

The auxiliary data is the patient status, which is the `lubomPD` vector that we extracted from the `Lubomski` data above. Similar to adding a metric, users can name their auxiliary data by specifying `name`. Note that users need to specify the evaluation metric to be used through the `metrics` argument, which has to match the evaluation metric added to the `TrioR` object.  

```{r}
# add auxiliary data to the Trio object
# name is the user-defined name, can be anything
# auxData is the value of the auxiliary data
# metric needs to be same name of the metric that has been added to the Trio object
trio$addAuxData(
  name = "patient_status", auxData = lubomPD, metrics = "Balanced Accuracy"
)
```

If users want to get evaluation metrics from the `TrioR` object, they can use the `getMetrics` function by referring the name of `auxData`.

```{r}
# get the metric from the Trio object
metrics <- trio$getMetrics("patient_status")
```

Step 3: Obtain relevant data for the model. To build a model, the following code will extract the data matrix and patient status outcome from `TrioR` object.   

```{r}
# get the gold standard from the Trio object
x <- trio$data
y <- trio$getAuxData("patient_status")
```

Step 4: For repeated cross-validation, the `split()` function will split the `y` vector (i.e., `auxData`) into the number of folds and repeats users want. In this case we used `n_fold = 2` and `n_repeat = 5` (i.e., 2-fold cross-validation with 10 repeats). Then users can get cross-validation indices (`cv_ind`) for each sample, through `splitIndices`. This gives a simple list where each element represents a combination of folds and repeats for each sample.  

```{r}
# get train and test indices
trio$split(y = y, n_fold = 2, n_repeat = 5)
cv_ind <- trio$splitIndices
```

Step 5: Build the classification model and evaluate. We can now use a for loop to cross-validate evaluation results.    

Using the `cv_ind`, user can subset to training and test data. As an example, we build a LASSO regression model as a classification model on the training data, then make predictions on the test data.   

Once we get the prediction, we pass the `pred` vector to `TrioR` and ref to the name of `auxData` (`patient_status` = `pred`). This tells the `evaluate()` function to compare the `pred` vector with auxData `patient_status` stored in the `TrioR` object. It will then computes the balanced accuracy, because this is the metric we have specified. 

```{r}
set.seed(1234)

result <- data.frame()

# loop through the 2 folds x 5 repeats = 10 runs
for (i in seq_along(cv_ind)) {
  train_id <- cv_ind[[i]]

  x_train <- x[train_id, ]
  x_test <- x[-train_id, ]
  y_train <- y[train_id]
  y_test <- y[-train_id]

  #  find the best lambda for LASSO regression
  cv_lasso <- cv.glmnet(as.matrix(x_train), y_train,
    alpha = 1, family = "binomial"
  )
  lam <- cv_lasso$lambda.1se

  # fit a model with the best lambda on training data
  fit <- glmnet(x_train, y_train, alpha = 1, lambda = lam, family = "binomial")

  # evaluate the model on test data
  pred <- predict(fit, x_test, s = "lambda.min", type = "class")
  pred <- as.factor(as.vector(pred))

  # get the chosen evaluation metric from the Trio
  eval_res <- trio$evaluate(
    list(lasso = list(patient_status = pred)),
    splitIndex = i
  )

  # keep track of the repeat and fold information
  eval_res$track <- names(cv_ind)[[i]]
  result <- rbind(result, eval_res)
}
```


After cross-validation, we can visualise cross-validation results by averaging results across folds within each repeats. 


```{r}
result$fold <- unlist(lapply(strsplit(result$track, ".", fixed = TRUE), `[`, 1))

result$repeats <- unlist(lapply(strsplit(result$track, ".", fixed = TRUE), `[`, 2))

result <- result %>%
  dplyr::group_by(datasetID, method, auxData, metric, repeats) %>%
  dplyr::summarize(result = mean(result))

# visualise the result
boxplot(result$result)

# look at the format of the result output
result
```


# Without cross validation 

In this example, we show a simpler case without cross validation, where we have a prediction that we want to compare with the ground truth directly. For example, we may have simulated a single-cell count matrix and want to compare with an experimental count matrix on various attributes. 

Here, we use the microbiome dataset to demonstrate the above case. 

Step 1: Simulate a matrix of the same size as the microbiome data using negative binomial.   
Step 2: Define a function that calculate the sparsity of a given matrix.      
Step 3: Define a metric that compare difference of two values.   
Step 4: Evaluate the difference in sparsity of the microbiome data and the simulated data.    

```{r}
set.seed(1)

# generate a simulated matrix
sim <- rnbinom(nrow(x) * ncol(x), size = 1, mu = 1)
sim <- Matrix(sim, nrow = nrow(x), ncol = ncol(x))

# function that calculate the sparsity of a matrix
calc_sparsity <- function(data) {
  sparsity <- sum(data == 0) / length(data)
  return(sparsity)
}


# metric that compare the difference of two values
calc_diff <- function(pred, gt) {
  return(gt - pred)
}
```

```{r}
# add metric that we just defined
trio$addMetric(name = "Difference", metric = calc_diff)

# calculate the sparsity of the data, and then input into the auxillary data
trio$addAuxData(name = "Sparsity", auxData = calc_sparsity(x), metrics = "Difference")


# because we used negative binomial to simulate a matrix
# we will name the method as "negative binomial"
# then calculate the sparsity of the simulated matrix and compare with the Sparsity auxiliary data
eval_res <- trio$evaluate(
  list(negative_binomial = list(Sparsity = calc_sparsity(sim)))
)

eval_res
```



From the evaluation result, we see there is a `r round(eval_res$result, 2)` difference in the sparsity between the microbiome data and the data simulated with negative binomial.   





# Importing result to BenchmarkInsights

The result dataframe obtained from Trio evaluation can be passed to BenchmarkInsights for subsequent visualisation. BenchmarkInsights is another structure in BenchHub that provide a list of functions to analyse and visualise the benchmarking results from multiple perspectives.

Note, to make the result dataframe compatible for BenchmarkInsights, please make sure the dataframe is in the format shown below, where the column names are "datasetID", "method", "auxData", "metric" , "result", and that there is one row for each result. 


```{r}
# in the with cross validation result, we need to average the results from multiple repeats to give one value
result <- result %>%
  dplyr::group_by(datasetID, method, auxData, metric) %>%
  dplyr::summarize(result = mean(result))

result <- rbind(result, eval_res)
result
```

Please see Vignette xxxxx for the details on the visualisations and functions in BenchmarkInsights.     


# Session Info

```{r}
sessionInfo()
```
