---
title: "5 BenchmarkStudy"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{5 Benchmark Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, warning=FALSE, message=FALSE, include=FALSE}
knitr::opts_chunk$set( warning = FALSE, message = FALSE)
library(R6)
library(scuttle) 
```

```{r}
library(BenchHub)
```

# Overview

The `BenchmarkStudy` object is designed to encapsulate all necessary components in a benchmarking study, including the data and functions associated. It provides a unified structure for benchmark developers to share their work and for method developers to interact with an existing benchmark study.

-   **Benchmark developers** can store *Trio* objects (containing the input data, metrics, and supporting evidence), any mapping functions and distribute a ready-to-use study object.\
-   **Method developers** can apply their methods to the provided data and evaluate their outputs using the built-in metrics.

This vignette provides a guide for both use cases under the current BenchHub submission workflow.

# For Benchmark Developer

This section demonstrates how to create a `BenchmarkStudy` object from a benchmarking study.

## Initialising the Study

We begin by creating an empty `BenchmarkStudy` object.

```{r}

study <- BenchmarkStudy$new()
 
```

```{r eval=FALSE}
# Download an existing Trio from the submission database
example_trio <- downloadSubmissionTrio("D001", cachePath = tempdir())

example_trio
```

## Define mapping function and protocol function

A mapping function is a helper function that processes method output into a format that can then be compared with the supporting evidence stored in the reference `Trio`. There are three ways to contribute the mapping function:

1.  Leave blank: if you don't want to contribute now
2.  Use existing GitHub repository: if your mapping function has been uploaded to the GitHub repository in the published paper
3.  Upload mapping functions stored in the study object
4.  Upload local mapping function to gist: define the mapping function locally

In this toy spatial transcriptomics example, the `Trio` contains the following supporting evidence:

-   `annotated_domain`
-   `celltype_proportions`

We therefore define two mapping functions that extract those objects from a method result.

Example 1: extract predicted spatial domains.

```{r}
# Define the mapping function 
extract_domains <- function(result) {
  if (is.data.frame(result) && "annotated_domain" %in% colnames(result)) {
    return(result$annotated_domain)
  }
  if (is.list(result) && "annotated_domain" %in% names(result)) {
    return(result$annotated_domain)
  }
  stop("Could not find 'annotated_domain' in the method output.")
}

# Add the mapping function
study$addMappingFunction(
  name = "annotated_domain",
  func = extract_domains,
  inputDescription = "Method output containing one predicted spatial domain label per spot.",
  outputDescription = "A vector of predicted spatial domain labels aligned to spots.",
  exampleUsage = paste(
    "## Minimal example",
    "#result <- list(annotated_domain = c('domain_1', 'domain_1', 'domain_2', 'domain_2'))",
    "#res <- study$runMapping('annotated_domain', result)",
    "#head(res)",
    sep = "\n"
  )
)
 
 
```

Example 2: extract predicted cell type proportions.

```{r}
# Define the mapping function 
extract_celltype_props <- function(result) {
  if (is.data.frame(result) && "celltype_proportions" %in% names(result)) {
    return(result$celltype_proportions)
  }
  if (is.list(result) && "celltype_proportions" %in% names(result)) {
    return(result$celltype_proportions)
  }
  if (is.matrix(result) || is.data.frame(result)) {
    mat <- as.matrix(result)
    rs <- rowSums(mat)
    rs[rs == 0] <- 1
    return(mat / rs)
  }
  stop("Could not extract cell type proportions from the method output.")
}

# Add the mapping function, it is optional but recommended to add example usage 
study$addMappingFunction(
  name = "celltype_proportions",
  func = extract_celltype_props,
  inputDescription = "Method output containing cell type proportions per spot.",
  outputDescription = "A matrix or data frame of cell type proportions aligned to spots.",
  exampleUsage = paste(
    "## Minimal example",
    "#props <- matrix(c(0.9, 0.1, 0.8, 0.2, 0.2, 0.8, 0.1, 0.9), ncol = 2, byrow = TRUE)",
    "#study$runMapping('celltype_proportions', props)",
    sep = "\n"
  )
)

```

Similar as mapping functions, the protocol function is the full workflow of benchmarking study. There are three ways to contribute the protocol function:

1.  Leave blank: if you don't want to contribute now
2.  Use existing protocol gist URL: if your protocol function has been uploaded to the GitHub repository in the published paper
3.  Upload local protocol file to gist

## Upload Study

Once the BenchmarkStudy object includes:

1.  A study name and description
2.  One or more Trio objects already represented in the submission database
3.  Mapping functions [optional]
4.  Protocol functions [optional]

the recommended next step is to an interactive console workflow via `interactivePrepareStudySubmission(study)`.

```{r eval=FALSE}
# Set name and description manually
study <- BenchmarkStudy$new(name = "ST toy study")
study$description <- "Toy spatial transcriptomics study."

interactivePrepareStudySubmission(study)
```

In that interactive workflow, BenchHub will guide you through:

-   selecting or confirming dataset IDs to link to the study,
-   entering the study description,
-   optionally providing a protocol gist,
-   optionally providing a mapping-functions gist or uploading mapping functions already stored in the `study` object,
-   reviewing the submission bundle, and
-   optionally submitting the Study immediately.

# For Method Developer

This section illustrates how a method developer can use the benchmark study object created by another user, apply their method, and evaluate its performance.

## Loading the Study

A `BenchmarkStudy` object can be downloaded from the submission database through its `studyID`.

```{r eval=FALSE}
loaded_study <- downloadSubmissionStudy(studyID = "ST005", cachePath = tempdir())
```

This returns a populated `BenchmarkStudy` object. For example:

```{r eval=FALSE}
loaded_study
loaded_study$name
loaded_study$description
loaded_study$version
length(loaded_study$trios)
```

Inspect the list of available trios, and available mapping functions

Each entry of `loaded_study$trios` is a `Trio` object with supporting evidence that can be used for evaluation.

```{r eval=FALSE}
length(loaded_study$trios)

loaded_study$trios[[1]]

```

This study provides mapping functions to process method outputs into a format that can be used for evaluation.

Each mapping function has documentation.

```{r eval=FALSE}
# list the names of the mapping function
loaded_study$listMappingFunctions()

# choose one to print the documentation 
loaded_study$printMappingFunctionDocumentation("annotated_domain")

```

## Preparing for evaluation

This benchmark study wants to assess predicted spatial domains and cell type proportions.

Suppose the method developer has run a method and obtained predicted domain labels and cell type proportions for each spot.

```{r eval=FALSE}
method_output <- list(
  annotated_domain = c("domain_1", "domain_1", "domain_2", "domain_2"),
  celltype_proportions = data.frame(
    celltype_A = c(0.9, 0.8, 0.2, 0.1),
    celltype_B = c(0.1, 0.2, 0.8, 0.9)
  )
)
```

The method developer can apply the mapping functions to the method output to generate the objects required for evaluation.

```{r eval=FALSE}

domain_pred <- loaded_study$runMapping("annotated_domain", method_output)
prop_pred <- loaded_study$runMapping("celltype_proportions", method_output)
```

## Evaluate

Now we can compare the simulated data against an experimental dataset using the `evaluate` function.

The evaluate function is in the format of `study$evaluate(trio_name, list(supporting evidence = output to compare with))`.

In the function below, the names in the list correspond to supporting evidence stored in the reference `Trio`.

```{r eval=FALSE}
 
result <- loaded_study$evaluate(loaded_study$trios[[1]]$name,  # name of the Trio to compare with
  list(
    "annotated_domain" = domain_pred,
    "celltype_proportions" = prop_pred
  ))

result

```

# Summary

This vignette demonstrated two ways that users can interact with the `BenchmarkStudy` framework:

-   Benchmark developers: create or update a `BenchmarkStudy` by adding `Trio` objects and optional mapping functions with clear documentation, then prepare and submit the study through the current Study submission workflow.

-   Method developers: load an existing `BenchmarkStudy` from the submission database, use the `Trio` objects to execute benchmarking methods of interest, use the mapping functions to convert method outputs where needed, and evaluate those outputs against the study’s supporting evidence using the `evaluate()` function.

# Session Info

```{r}

sessionInfo()

```
